---
title: "R Notebook"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: yes
  html_notebook:
    fig_height: 8
    fig_width: 12

---

# Phyloseq

## ggplot2 theme

```{R}
library(tidyverse)
library(stringr)
#### Theme for publication: http://areshenk-research-notes.com/publication-quality-plots-in-r/ ####

theme_pub <- function (base_size = 12, base_family = "") {
  theme_grey(base_size = base_size,
             base_family = base_family) %+replace%

    theme(# Set text size
      plot.title = element_text(size = 18),
      axis.title.x = element_text(size = 16),
      axis.title.y = element_text(size = 16,
                                  angle = 90),
      axis.text.x = element_text(size = 14,colour="black",element_text(margin = margin(5,0,0,0),
                                                                       angle = 90)),
      axis.text.y = element_text(size = 14,colour="black",element_text(margin = margin(0,5,0,0))),
      strip.text.y = element_text(size = 15,
                                  angle = -90),

      # Legend text
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 15),
      legend.background = element_rect(fill = "transparent", colour=NA),

      # Configure lines and axes
      axis.ticks.x = element_line(colour = "black"),
      axis.ticks.y = element_line(colour = "black"),

      # Plot background
      panel.background = element_rect(fill = "transparent",colour=NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill="transparent", colour=NA),

      # Facet labels       
      legend.key = element_blank(),
      strip.background = element_rect(fill = "transparent",
                                      colour = NA,
                                      size = 0.1),
       panel.border = element_rect(colour = "black", fill="transparent"),
       strip.text.x = element_text(size=18,face="bold",margin=margin(0,0,5,0))
      )

}

# Create colour palette for taxonomy plotting

colours <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
             "#e31a1c","#fdbf6f","#ff7f00","#cab2d6","#6a3d9a",
             "#ffff99","#b15928","#8BC395","#919D5F","#EB494A",
             "#F79C5D","#DEB969")

```

Other useful function

```{R}
# Figure out highest unclassified rank of an OTU/DSV
name_otus = function(ps) {
  
  otuName = as.character(tax_table(ps)[,"genus"])
  otuName = str_split_fixed(otuName, "_", 2)[,1]
  names(otuName) = taxa_names(ps)
  
  for (level in c("family", "order", "class", "phylum","kingdom")) {
    uncl = (otuName=="")
    if (!any(uncl)) break
    otuName[uncl] = as.character(tax_table(ps)[uncl,level])
    otuName[uncl] = str_split_fixed(otuName[uncl], "\\(", 2)[,1]
  }
  
  otuName = sub("_[1-9]$", "", otuName) # remove appendices like "_1"
  otuName = factor(otuName, unique(otuName))
  otuName = otuName[order(otuName)]
  
  return(otuName)
}

# Create table for taxonomy plotting 
taxplottable = function (ps) {
  
  # Add new column with highest unclassified taxonomic rank to ps object
  highest_unclassified_rank <- name_otus(ps) %>%
    data.frame(OTU=names(.), highest_unclassified_rank = .) %>%
    mutate(highest_unclassified_rank = as.character(highest_unclassified_rank))
  taxtable <- as.data.frame(tax_table(ps), stringsAsFactors=F) %>%
    mutate(OTU=row.names(tax_table(ps))) %>%
    left_join(highest_unclassified_rank) %>%
    select(-OTU)
  row.names(taxtable) <- row.names(tax_table(ps))
  tax_table(ps) <- as.matrix(taxtable)
  
  # Figure out top12 genera
  ps_genus <- tax_glom(ps, taxrank="highest_unclassified_rank") %>% # merge by highest_unclassified_rank
  transform_sample_counts(.,  function(OTU) OTU/sum(OTU)) # calculate relabun
  top12_otus <- names(sort(taxa_sums(ps_genus),TRUE)[1:12])
  top12_genera <- as.character(tax_table(ps_genus)[top12_otus,"highest_unclassified_rank"]) # get names
  
  # Calculate relative abundance
  ps_relabun <- transform_sample_counts(ps,  function(OTU) OTU/sum(OTU))

  # Convert phyloseq object into large dataframe and manipulate
  ps_relabun_df <- psmelt(ps_relabun) %>% 
    group_by(OTU, Fermentation)  %>%
    filter(max(Abundance) > 0.01) %>% # filter out low abundant OTUs
    ungroup %>%
    filter(highest_unclassified_rank %in% top12_genera) # keep only top 12 genera
     

  return(ps_relabun_df)
}

# Create distance matrix and reformat for easy plotting
calculateDistMat <- function(phylo_obj){
  # Calculate distance matrix
  dm <- vegdist(otu_table(phylo_obj), method="bray")
  
  # Reformat the  to long format matrix
  m <- data.frame(t(combn(rownames(otu_table(phylo_obj)),2)), as.numeric(dm))
  names(m) <- c("sample1", "sample2", "distance")

  # Add factor 'Day'
  m$day_sample1 <- str_split_fixed(m$sample1,"_",4)[,2]
  m$day_sample2 <- str_split_fixed(m$sample2,"_",4)[,2]
  
  # change Day 28 to 30
  m$day_sample1 <- gsub("28","30",m$day_sample1)
  m$day_sample2 <- gsub("28","30",m$day_sample2)

  # Keep only samples from same day
  m <- m[which(m$day_sample1== m$day_sample2),]
  
  return(m)
}  
```

## Load in data

```{R}
library(phyloseq); packageVersion("phyloseq")
load("../intermediate/ps.Robject")
ps
```

## Create sample data from sample name and subset to CJ4 only

```{R}
# Make a data.frame holding the sample data
sample_data <- sample_data(ps) %>%
  mutate(rowNames = row.names(sample_data(ps)),
         Fermentation = str_sub(description, 1, 3)) %>%
  filter(Fermentation %in% c("CJ4")) %>%
  mutate(Day = str_sub(description, 5, 6),
         Repeat = str_sub(description, 7, 7),
         Source = str_sub(description, 8, -1)) %>%
  mutate(Source = str_replace(Source, "gDNA", "DNA")) %>%
  mutate(Source = str_replace(Source, "cDNA", "RNA"))
  
# Subset phyloseq object
row.names(sample_data) <- sample_data$rowNames
ps <- phyloseq(otu_table(ps),
               sample_data(sample_data),
               tax_table(ps))
ps
```


## Remove the bad samples

```{R}
ps <- prune_samples(sample_sums(ps) > 0,ps)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
```


# Abundance prior to chloroplast removal

```{r}
ps_melt <- psmelt(ps)
prefilter_abun <- ps_melt %>%
  group_by(Source, description) %>%
  summarise(TotalCount = sum(Abundance)) %>%
  mutate(Day = str_sub(description, 5,6), filter = "NO")
prefilter_abun
```

## Remove Chloroplast and Mitochondria

```{r}
# non_bacterial
taxtable <- as.data.frame(tax_table(ps)) 
non_bacterial <- taxtable$kingdom!="Bacteria" | is.na(taxtable$kingdom)
ps <- prune_taxa(!non_bacterial,ps)

# Chloroplast and mitochondria
taxtable <- as.data.frame(tax_table(ps)) 
chloroplast_mitochondria <- taxtable$class=="Chloroplast" | taxtable$family=="Mitochondria"


# Remove
ps <- prune_taxa(!chloroplast_mitochondria,ps)

ps
```

# Abundance after chloroplast removal

```{r}
ps_melt <- psmelt(ps)
postfilter_abun <- ps_melt %>%
  group_by(Source, description) %>%
  summarise(TotalCount = sum(Abundance)) %>%
  mutate(Day = str_sub(description, 5,6), filter = "YES")
postfilter_abun
```
```{r}
plot <- rbind(prefilter_abun, postfilter_abun) %>%
  rename(Abundance = TotalCount) %>%
  spread(key = filter, value= Abundance) %>%
  mutate(diffAbundance = NO - YES) %>%
  select(-NO) %>%
  rename(Bacterial = YES, NonBacterial = diffAbundance) %>%
  gather(key = "Origin", value = "Abundance", Bacterial, NonBacterial) %>%
  mutate(Repeat = str_sub(description, 7, 7), 
         Day = str_c("Day ", Day, sep = '')) %>%
  ggplot(aes(x = Repeat, y = Abundance, fill = Origin)) +
    geom_bar(stat = "identity", 
             color = "black",
             position = position_fill(reverse = T)) + 
    ylab("Relative abundance (%) \n") + 
    xlab("") +
    theme_pub() +
    labs(fill="") + 
    facet_grid(Source ~ Day, drop=T) +
    scale_fill_manual(values = colours) +
    theme(axis.text.x= element_blank(), 
          axis.ticks.x=element_blank(), 
          legend.position="right",
          strip.text.y = element_text(size = 15, face = "bold", margin = margin(0,0,0,4)))
  
plot

ggsave("~/serverUA/sander/DADA2_WannesSander_170821/analysis/Nonbacterialreads.svg", plot = plot)
saveRDS(plot, "~/serverUA/sander/DADA2_WannesSander_170821/analysis/Nonbacterialreads.rds")
```

## Plot Alpha div

### Shannon

```{R}
title <- "Alpha Diversity \n"
p <- plot_richness(ps, measures=c("Observed","Shannon"), x='Day')

# Remove layer of points to plot over jitter points
p$layers <- p$layers[-1]

p + geom_point(size = 2, position= position_jitterdodge(),aes(colour=Source)) +
  theme_pub() +
  theme(legend.position= 'right',axis.text.x=element_text(margin=margin(5,0,0,0),angle=45,hjust=1, vjust=1)) +
  ggtitle(title) +
  ylab("Alpha diversity measure \n") +
  xlab("\n Day") +
  expand_limits(y=0)
```

## Calculate Relative Abundance

```{R}
ps_relabun <- transform_sample_counts(ps,  function(OTU) OTU/sum(OTU))
```

## Betadiv

```{R}
library(ggrepel)
title <- "PCoA (Bray-Curtis distance) \n"
ordination <- ordinate(ps_relabun,"PCoA","bray")
p <- plot_ordination(ps_relabun, ordination, color="Source")
p <- p + geom_point(size=3) + ggtitle(title) +
  geom_text_repel(aes(label = Day))
p
```

## Heatmap

```{r }
top50taxa <- prune_taxa(names(sort(taxa_sums(ps_relabun),TRUE)[1:50]), ps_relabun)
plot_heatmap(top50taxa,title="OTU heatmap: Top 50 taxa","Genus") + theme(axis.text.y=element_blank()) +
  facet_wrap(~Source, scales = 'free_x')
```

## Taxonomic plots

```{r}
# Calculate relative abundance and prepare data for plotting
plottable <- taxplottable(ps)
```

```{r}
# Plot using ggplot2
ggplot(plottable, aes(x=Repeat, y=Abundance, fill=highest_unclassified_rank)) + geom_bar(stat = "identity", color = "black") + 
  ylab("Percentage of Sequences \n") + 
  xlab("") +
  theme_pub() +
  theme(axis.text.x= element_blank(), axis.ticks.x=element_blank(), legend.position="bottom") +
  labs(fill="") + 
  facet_grid(Source ~ Day,scales="free",drop=T) +
  scale_fill_manual(values = colours) +
  scale_y_continuous(expand=c(0,0)) + 
  guides(fill=guide_legend(byrow=F,ncol=2)) 
```

One sample shows inconsistency with the rest of the samlpes (RNA, DAY 10). But we can still average everything out. 

Let’s calculate their mean abundance per day per OTU per fermentation and start parsing the data for plotting.  

```{r}
# Caluclate mean abundance per OTU per Day per fermentation
plottable_mean <- plottable %>%
  group_by(OTU, Day, Source) %>%
  summarise(meanAbundance= mean(Abundance))

plottable_mean_metadata <- plottable %>%
  select(-Repeat,-Sample, -Abundance, - rowNames, -description)%>%
  left_join(plottable_mean) %>%
  distinct() %>%
  mutate(Day = as.numeric(as.character(Day)), meanAbundance = meanAbundance * 100)  # Change Day to numeric and multiply abundance by 100

# Relevel highest_unclassified_rank to match colours of main genera with other plots
common_taxa <- c("Leuconostoc", "Lactobacillus", "Lactococcus", "Enterobacteriaceae", "Citrobacter", "Klebsiella", "Pantoea", "Raoultella") 
additional_taxa <- plottable_mean_metadata$highest_unclassified_rank %>%
  unique() %>%
  .[!. %in% common_taxa] %>%
  as.character()

Levels <- c(common_taxa,additional_taxa)

plottable_mean_metadata$highest_unclassified_rank <- factor(plottable_mean_metadata$highest_unclassified_rank, levels=Levels)
```

```{r, fig.width=12, fig.height=8}
plot <- ggplot(plottable_mean_metadata, aes(x=Day, y=meanAbundance, group=OTU)) + 
  xlab("\n Day") +
  ylab("Relative abundance\n") +
  geom_area(aes(fill=highest_unclassified_rank, group=OTU),colour="white", size=0.2) + 
  geom_vline(aes(xintercept=Day),colour='black', linetype='dotted', alpha= 0.8) +
  theme_pub() + 
  scale_fill_manual(values=colours[-4]) +
  theme(legend.title=element_blank(), legend.text = element_text(face ="italic"),
        panel.border = element_blank(),
        axis.text.x= element_text(size=10, margin=margin(2,0,0,0)),
        axis.text.y= element_text(size=10, margin=margin(0,0,0,2)),
        legend.position="right") +
  expand_limits(y=0,x=0) + 
  scale_y_continuous(limits=c(0,100),breaks=c(0,50,100)) +
  scale_x_continuous(breaks=c(0,10,20,30,40,50,60)) +
  theme(plot.margin=unit(c(0,1,0,0),"cm")) + 
  guides(fill=guide_legend(byrow = F,ncol=1)) + 
  facet_wrap(~Source, nrow = 2)

plot

ggsave("~/serverUA/sander/DADA2_WannesSander_170821/analysis/Tax_comp.svg", last_plot())
saveRDS(plot, "~/serverUA/sander/DADA2_WannesSander_170821/analysis/tax_comp.rds")

```

```{r, fig.width=12, fig.height=8}
ggplot(plottable_mean_metadata, aes(x=Day, y=meanAbundance, group=OTU)) + 
  xlab("\n Day") +
  ylab("Relative abundance\n") +
  geom_area(aes(fill=order, group=OTU),colour="white", size=0.2) + 
  geom_vline(aes(xintercept=Day),colour='black', linetype='dotted', alpha= 0.8) +
  theme_pub() + 
  scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02')) +
  theme(legend.title=element_blank(), legend.text = element_text(face ="italic"),
        panel.border = element_blank(),
        axis.text.x= element_text(size=10, margin=margin(2,0,0,0)),
        axis.text.y= element_text(size=10, margin=margin(0,0,0,2)),
        legend.position="right") +
  expand_limits(y=0,x=0) + 
  scale_y_continuous(limits=c(0,100),breaks=c(0,50,100)) +
  scale_x_continuous(breaks=c(0,10,20,30,40,50,60)) +
  theme(plot.margin=unit(c(0,1,0,0),"cm")) + 
  guides(fill=guide_legend(byrow = F,ncol=1)) + 
  facet_wrap(~Source, nrow = 2)

```

```{r}
lactos <- plottable %>%
  filter(genus == "Lactobacillus") %>%
  mutate(species_wrap = str_wrap(species),width = 55)
  
# Plot using ggplot2
ggplot(lactos, aes(x=Repeat, y=Abundance, fill=species_wrap)) + geom_bar(stat = "identity", color = "black") + 
  ylab("Percentage of Sequences \n") + 
  xlab("") +
  theme_pub() +
  theme(axis.text.x= element_blank(), axis.ticks.x=element_blank(), legend.position="bottom",
        legend.text = element_text(size = 9)) +
  labs(fill="") + 
  facet_grid(Source ~ Day,scales="free",drop=T) +
  scale_fill_manual(values = colours) +
  scale_y_continuous(expand=c(0,0)) + 
  guides(fill=guide_legend(byrow=F,ncol=2)) 
```

# Some stats for paper

```{r}
plottable_mean_metadata %>%
  filter(genus == "Lactobacillus") %>%
  group_by(Source, Day) %>%
  summarise(maxAbun = max(sum(meanAbundance)), numberASVs = n())

```
```{r}
plottable_mean_metadata %>%
  filter(genus == "Leuconostoc") %>%
  group_by(Source, Day) %>%
  summarise(maxAbun = max(sum(meanAbundance)), numberASVs = n())

```

```{r}
plottable_mean_metadata %>%
  filter(genus == "Lactococcus") %>%
  group_by(Source, Day) %>% 
  summarise(maxAbun = max(sum(meanAbundance)), numberASVs = n())

```

```{r}
plottable_mean_metadata %>%
  filter(family == "Enterobacteriaceae") %>%
  group_by(Source, Day) %>% 
  summarise(maxAbun = max(sum(meanAbundance)), numberASVs = n())

```
